---
title: "JSM tutorial on genetic fine-mapping with summary statistics: the RSS likelihood and the SuSiE model"
author: Peter Carbonetto
output: workflowr::wflow_html
---

The goal of this tutorial is to give some intuition for the key
statistical concepts and computations involved in genetic fine-mapping
with the [SuSiE][susie] and [SuSiE-RSS][susie_rss] models without
entering too deeply into the mathematical derivations.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Initial setup
-------------

For this tutorial, I'm using a lightly modified version of susieR
available in the "output_residuals" branch:

```{r install-susie, eval=FALSE}
library(remotes)
install_github("stephenslab/susieR@output_residuals")
```

Load all the packages used in the tutorial:

```{r load-pkgs}
library(ggplot2)
library(cowplot)
library(susieR)
packageVersion("susieR")
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

The data set contains genotype and quantitative trait data for $n
= 574$ individuals and $p = 750$ genetic markers (SNPs).

```{r load-data}
load("../data/jsm_tutorial_data.RData")
n <- nrow(X)
dim(X)
```

Association analysis vs. fine-mapping 
-------------------------------------

First let's connect a genetic association analysis to a fine-mapping
analysis in order to understand points of overlap and their
differences.

This is a function we will use in the code chunks below to create a
basic Manhattan plot. Understanding how this function works is not
important.

```{r manhattan-plot}
manhattan_plot <- function (pval, causal_snps = NULL, xlab = "") {
  n <- length(pval)
  dat <- data.frame(pos = 1:n,pval = -log10(pval))
  out <- ggplot(dat,aes(x = pos,y = pval)) +
    geom_point() +
	scale_x_continuous(breaks = NULL) +
	labs(x = xlab,y = "-log10 p-value") +
    theme_cowplot(font_size = 12)
  if (!is.null(causal_snps)) {
	dat_causal <- dat[causal_snps,]
    out <- out + geom_point(data = dat_causal,color = "red")
  }
  return(out)
}
```

The Manhattan plot is a visualization of the association analysis: it
summarizes the evidence for association with the quantitative trait at
the 740 SNPs.

```{r show-manhattan-plot, fig.height=1.75, fig.width=5}
manhattan_plot(pval,xlab = "SNPs on human chromosome 1 near the FMO2 gene")
```

There are several "clusters" of SNPs that have very low
*p*-values. This clustering is typical due to strong correlations
("LD") between nearby SNPs. The fact that there are several clusters
suggests that perhaps there is more than one SNP affecting the
quantitative trait, but it is hard to say for sure based on an
association analysis alone.

Let's now run a SuSiE fine-mapping analysis *using the same statistics
that were used in the association tests*—specifically, we will use the
z-scores. (We will also eventually need ${\bf R}$, the $p \times p$
matrix of correlations between all SNPs, also sometimes called the "LD
matrix".)

```{r susie-rss, fig.height=1.75, fig.width=5}
fit <- susie_rss(zhat,R,n = n,estimate_prior_method = "EM",
                 tol = 1e-8,min_abs_corr = 0)
```

This is a function we will use in the code chunks below to visualize
the results of the SuSiE fine-mapping analysis. Understanding how this
function works is not important.

```{r susie-plot-better}
pip_plot <- function (fit, causal_snps = NULL, xlab = "") {
  cs_colors <- c("dodgerblue","limegreen")
  pip <- fit$pip
  n   <- length(pip)
  dat <- data.frame(pos = 1:n,pip = pip,CS = as.character(NA))
  num_cs <- length(fit$sets$cs)
  for (i in 1:num_cs) {
    j <- fit$sets$cs[[i]]
	dat[j,"CS"] <- paste0("CS",i)
  }
  dat_cs <- subset(dat,!is.na(CS))
  out <- ggplot(dat,aes(x = pos,y = pip)) +
    geom_point() +
	geom_point(data = dat_cs,shape = 1,size = 3,
	           mapping = aes(color = CS)) +
	scale_x_continuous(breaks = NULL) +
	scale_color_manual(values = cs_colors) + 
	labs(x = xlab,y = "PIP") +
    theme_cowplot(font_size = 12)
  if (!is.null(causal_snps)) {
	dat_causal <- dat[causal_snps,]
    out <- out + geom_point(data = dat_causal,color = "red")
  }
  return(out)
}
```

One thing we did not show in the previous plot is that there are
actually 2 "causal SNPs"—that is, 2 SNPs that affect the trait, shown
as 2 red dots in the next plot. (This is a simulated quantitative
trait, so we know the truth.)

```{r susie-rss-plot, fig.height=1.75, fig.width=5.25}
causal_snps <- which(b_true != 0)
pip_plot(fit,causal_snps,
         xlab = "SNPs on human chromosome 1 near the FMO2 gene")
```

This plot is a "PIP plot", where PIP = "posterior inclusion
probability". The PIPs weigh the support a SNP being a causal SNP, and
indeed, the two causal SNPs have the highest PIPs. Each of the two
credible sets (CSs) is meant to capture a causal SNP with high
probability, and indeed in this example they do. And then the PIPs
within each CS capture the uncertainty in which of the nearby SNPs is
the true causal SNP. In CS2, several SNPs are strongly correlated with
each other—the minimum correlation among the four SNPs in CS2
is 0.99—and so it difficult to say for sure which of these SNPs is the
causal SNP. The PIPs in CS2 reflect this. 

```{r cs2-correlations}
j <- fit$sets$cs$L1
min(R[j,j])
```

By contrast, in CS1 we are quite confident (PIP > 0.95) that the one
SNP is causal.

To reinforce some of these points, it is helpful to compare the
association analysis directly with the fine-mapping analysis:

```{r more-susie-plots, fig.height=3.25, fig.width=5.25}
p1 <- manhattan_plot(pval,causal_snps,
                     xlab = "SNPs on human chromosome 1 near the FMO2 gene")
p2 <- pip_plot(fit,causal_snps,
               xlab = "SNPs on human chromosome 1 near the FMO2 gene")
plot_grid(p1,p2,nrow = 2,ncol = 1,align = "v",axis = "lr")
```

The special case of 1 causal SNP
--------------------------------

A goal of this tutorial is to gain some insight into how we arrived at
this fine-mapping result and how SuSiE works. Starting with the case
of a single causal SNP will be our first step toward that goal. 

Assuming a single causal SNP simplifies things greatly because we can
compute the PIPs and CS quite easily using the z-scores. Here we break
down the analysis into three calculations: the calculation of the CS
given the PIPs; the calculation of the PIPs given the Bayes factors
(BFs); and the calculation of the BFs.

Computing the PIPs given the BFs is quite simple. If we assume the
prior inclusion probabilities are the same for all SNPs, we have
$$
\mathrm{PIP}_j = \frac{\mathrm{BF}_j}{\sum_{j'=1}^p \mathrm{BF}_{j'}}.
$$
This function performs this calculation in a numerically stable way
given the log-BFs:

```{r normalizelogweights}
normalizelogweights <- function (logw) {
  x <- max(logw)
  w <- exp(logw - x)
  return(w/sum(w))
}
```

This function returns a CS given the PIPs:

```{r get-cs}
get_cs <- function (pip, coverage = 0.95) {
  i <- order(pip,decreasing = TRUE)
  n <- which(cumsum(pip[i]) > coverage)[1]
  return(i[1:n])
}
```

The Bayes factor for each SNP $j$ is a ratio of two likelihoods
comparing a "single-SNP model" of the quantitative trait
$\boldsymbol{y}$ against a "null model" in which no SNP affects the
trait. In most of our analyses the single-SNP model for each SNP $j$
is a simple linear regression in which the prior distribution for the
unknown coefficient $b_j$ is a normal distribution centered at zero
(for simplicity I'll assume the residual variance is 1, which is an
assumption we will make later on anyhow):
$$
\begin{aligned}
\boldsymbol{y} \mid b_j &\sim
N_n(\boldsymbol{x}_j b_j, 1) \\
b_j \mid \sigma_0^2 &\sim N(0, \sigma_0^2)
\end{aligned}
$$

A key idea is this single-SNP model can be reframed as a model of the
z-score obtained from the association test:
$$
\begin{aligned}
z_j \sim N(\sqrt{n} b_j, 1) \\
b_j \sim N(0, \sigma_0^2).
\end{aligned}
$$
Therefore, the BF calculation can be expressed as a comparison of two
models of the z-scores:
$$
\begin{aligned}
\mathrm{BF}_j &=
\frac{p(\boldsymbol{y} \mid \boldsymbol{x}_j, b_j \neq 0)}
     {p(\boldsymbol{y} \mid \boldsymbol{x}_j, b_j = 0)} \\
&= \frac{p(z_j \mid b_j \neq 0)}
        {p(z_j \mid b_j = 0)} \\
&\propto \exp\bigg\{
\frac{z_j^2}{2} \times
\frac{\tilde{\sigma}_0^2}{\tilde{\sigma}_0^2 + 1}
\bigg\}, \\
\end{aligned}
$$
where $\tilde{\sigma}_0^2 = n \sigma_0^2$.  The following function
implements the BF calculation for all SNPs. (It also computes the
posterior mean effect estimates $E[b_j \mid z_j, \sigma_0^2]$ which we
do not need immediately, but we will need later on.)

```{r compute-log-bf}
compute_log_bf <- function (z, s0) {
  s0  <- n * s0
  lbf <- z^2/2 * s0/(s0 + 1)
  b1  <- z/(1 + 1/s0)
  return(list(lbf = lbf,b1 = b1))
}
```

Compare to eq. A.3 of the [SuSiE paper][susie]  and eq. 17 of
the [SuSiE-RSS paper][susie_rss].

Let's now perform these calculations on the example data set:

```{r single-snp, fig.height=1.75, fig.width=5.25}
s0  <- 0.36
z   <- sqrt(n/(n + zhat^2)) * zhat
lbf <- compute_log_bf(z,s0)$lbf
pip <- normalizelogweights(lbf)
cs <- get_cs(pip)
pip_plot(list(pip = pip,sets = list(cs = list(L1 = cs))),causal_snps,
         xlab = "SNPs on human chromosome 1 near the FMO2 gene")
```

Note that we adjusted the z-scores in the same way as in
`susie_rss()`. See the SuSiE-RSS paper for an explanation of why this
is done.

(Note that the prior variance setting of 0.36 here was mainly chosen
to match the estimate obtained by running `susie_rss()`, and the
particular choice is not critical.)

Not surprisingly, because we assumed 1 causal SNP, the CS captures
only one of the two causal SNPs. Despite this failure, we still made
some progress by reusing the association tests to perform fine-mapping
(that is, we computed a CS and a PIP for each SNP).

A quick verification that these calculations match the `susie_rss()`
result:

```{r, fig.height=3, fig.width=3}
par(mar = c(4,4,2,1))
fit_1cs <- susie_rss(zhat,R,n = n,L = 1,estimate_prior_method = "EM",
                     tol = 1e-6)
plot(fit_1cs$pip,pip,pch = 20,xlab = "susie_rss PIP",ylab = "JSM tutorial PIP")
abline(a = 0,b = 1,lty = "dotted",col = "magenta")
```

Extending to more than one causal SNP: a simple iterative calculation
---------------------------------------------------------------------

ADD TEXT HERE.

```{r ibss}
ibss <- function (z, R, n, L, s0, numiter = 10, coverage = 0.95) {
  p   <- length(zhat)
  PIP <- matrix(0,p,L)
  B1  <- matrix(0,p,L)
  r   <- z
  residuals <- matrix(0,p,L)
  for (iter in 1:numiter) {
    for (l in 1:L) {
      b1  <- B1[,l]
      r   <- r + R %*% b1
	  residuals[,l] <- r
	  out <- compute_log_bf(r,s0[l])
	  pip <- normalizelogweights(out$lbf)
	  b1  <- out$b1 * pip
      r   <- r - R %*% b1
	  PIP[,l] <- pip
	  B1[,l]  <- b1
    }
  }
  return(list(pip = rowSums(PIP),
              residuals = residuals,
              sets = list(cs = apply(PIP,2,function (x) get_cs(x,coverage)))))
}
res <- ibss(z,R,n,L = 2,s0 = c(0.3,0.2),numiter = 10)
p1 <- pip_plot(res,causal_snps,
               xlab = "SNPs on human chromosome 1 near the FMO2 gene")
range(res$pip - fit$pip)
```

```{r, eval=FALSE}
res_iter1 <- ibss(z,R,n,L = 2,s0 = c(0.3,0.2),numiter = 1)
p1 <- pip_plot(res1,causal_snps,
               xlab = "SNPs on human chromosome 1 near the FMO2 gene")
res_iter10 <- ibss(z,R,n,L = 2,s0 = c(0.3,0.2),numiter = 10)
p2 <- pip_plot(res10,causal_snps,
               xlab = "SNPs on human chromosome 1 near the FMO2 gene")
plot_grid(p1,p2,nrow = 2,ncol = 1)
qplot(z,res1$residuals[,2],color = abs(R[,679]))
```

Multi-SNP model:
$$
\begin{aligned}
\boldsymbol{y} \mid \boldsymbol{b} &\sim
N_n({\bf X}{\boldsymbol b}, {\bf I}_n) \\
\boldsymbol{b} \mid \sigma_0^2 &\sim N_p(0, \sigma_0^2 {\bf I}_p).
\end{aligned}
$$

Multi-SNP regression likelihood:
$$
{\boldsymbol y} \mid {\boldsymbol b}, \sigma^2 \sim
N_n({\bf X}{\boldsymbol b}, \sigma^2 {\bf I}_n)
$$
Equivalent likelihood formulated using summary statistics:
$$
{\boldsymbol t} \mid {\bf R}, n, \tilde{\boldsymbol b}, \sigma^2 \sim
N_p(\sqrt{n} {\bf R} \tilde{\boldsymbol b}, \sigma^2{\bf R})
$$
With a few additional assumptions, the
RSS likelihood using z-scores (actually these are "lightly modified"
z-scores):
$$
\tilde{\boldsymbol z} \mid {\bf R}, n, {\boldsymbol b}, \sigma^2 \sim
N_p(\sqrt{n} {\bf R} \boldsymbol b, \sigma^2{\bf R})
$$

${\bf R}$ is the $p \times p$ correlation ("LD") matrix;
${\boldsymbol z}$ is a "lightly modified" z-score. Refer to the corresponding
equation for the RSS likelihood in the SuSiE-RSS paper.

NEXT: Implement a simple version of the IBSS algorithm using the
RSS likelihood. See Algorithm 1 in the S1 Text of the SuSiE-RSS paper.

### References

Wang et al (2020). A simple new approach to variable selection in
regression, with application to genetic fine mapping. *Journal of the
Royal Statistical Society, Series B* 82, 1273-1300.
[link][susie]

Zou et al (2022). Fine-mapping from summary data with the Sum of
Single Effects model. *PLoS Genetics* 18, e1010299.
[link]

[susie]: https://doi.org/10.1101/501114
[susie_rss]: https://doi.org/10.1371/journal.pgen.1010299
