---
title: "JSM tutorial on fine-mapping with summary statistics: the RSS likelihood and the SuSiE model"
author: Peter Carbonetto
output: workflowr::wflow_html
---

ADD SUMMARY OF THE TUTORIAL HERE.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Initial setup
-------------

For this tutorial, I'm using a lightly modified version of susieR in
the "output_residuals" branch:

```{r install-susie, eval=FALSE}
library(remotes)
install_github("stephenslab/susieR@output_residuals")
```

Load all the packages used in the tutorial:

```{r load-pkgs}
library(ggplot2)
library(cowplot)
library(susieR)
packageVersion("susieR")
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

The data set contains genotype and quantitative trait data for $n
= 574$ individuals and $p = 750$ genetic markers (SNPs).

```{r load-data}
load("../data/jsm_tutorial_data.RData")
n <- nrow(X)
dim(X)
```

Association analysis vs. fine-mapping 
-------------------------------------

First let's relate a typical genetic association analysis to a
fine-mapping analysis, and understand the points of overlap and their
differences.

This is a function we will use to create a basic Manhattan plot:

```{r manhattan-plot}
manhattan_plot <- function (pval, causal_snps = NULL, xlab = "") {
  n <- length(pval)
  dat <- data.frame(pos = 1:n,pval = -log10(pval))
  out <- ggplot(dat,aes(x = pos,y = pval)) +
    geom_point() +
	scale_x_continuous(breaks = NULL) +
	labs(x = xlab,y = "-log10 p-value") +
    theme_cowplot(font_size = 12)
  if (!is.null(causal_snps)) {
	dat_causal <- dat[causal_snps,]
    out <- out + geom_point(data = dat_causal,color = "red")
  }
  return(out)
}
```

The Manhattan plot is a visualization of the association analysis: it
summarizes the evidence for association with the quantitative trait at
each of the 740 SNPs.

```{r show-manhattan-plot, fig.height=1.75, fig.width=5}
manhattan_plot(pval,xlab = "SNPs on human chromosome 1 near the FMO2 gene")
```

ADD TEXT HERE.

```{r susie-rss, fig.height=1.75, fig.width=5}
fit <- susie_rss(zhat,R,n = n,estimate_prior_method = "EM",
                 tol = 1e-6,min_abs_corr = 0)
```

ADD TEXT HERE.

```{r susie-plot-better}
susie_plot_better <- function (fit, causal_snps = NULL, xlab = "") {
  cs_colors <- c("dodgerblue","limegreen")
  pip <- fit$pip
  n   <- length(pip)
  dat <- data.frame(pos = 1:n,pip = pip,CS = as.character(NA))
  num_cs <- length(fit$sets$cs)
  for (i in 1:num_cs) {
    j <- fit$sets$cs[[i]]
	dat[j,"CS"] <- paste0("CS",i)
  }
  dat_cs <- subset(dat,!is.na(CS))
  out <- ggplot(dat,aes(x = pos,y = pip)) +
    geom_point() +
	geom_point(data = dat_cs,shape = 1,size = 3,
	           mapping = aes(color = CS)) +
	scale_x_continuous(breaks = NULL) +
	scale_color_manual(values = cs_colors) + 
	labs(x = xlab,y = "PIP") +
    theme_cowplot(font_size = 12)
  if (!is.null(causal_snps)) {
	dat_causal <- dat[causal_snps,]
    out <- out + geom_point(data = dat_causal,color = "red")
  }
  return(out)
}
```

ADD TEXT HERE.

```{r susie-rss-plot, fig.height=1.75, fig.width=5.25}
causal_snps <- which(b_true != 0)
susie_plot_better(fit,causal_snps,
                  xlab = "SNPs on human chromosome 1 near the FMO2 gene")
```

ADD TEXT HERE.

```{r more-susie-plots, fig.height=3.25, fig.width=5.25}
p1 <- manhattan_plot(pval,causal_snps,
                     xlab = "SNPs on human chromosome 1 near the FMO2 gene")
p2 <- susie_plot_better(fit,causal_snps,
                        xlab = "SNPs on human chromosome 1 near the FMO2 gene")
plot_grid(p1,p2,nrow = 2,ncol = 1,align = "v",axis = "lr")
```

ADD TEXT HERE.

```{r}
normalizelogweights <- function (logw) {
  x <- max(logw)
  w <- exp(logw - x)
  return(w/sum(w))
}
```

```{r}
compute_log_bayes_factor <- function () {
}
```

TO DO:

+ Give expression for the Bayes factor.

+ Wrap the BF calculation into a function.

See eq. A.3 of the SuSiE paper and eq. 17 of the SuSiE-RSS paper.


```{r}
z   <- sqrt(n/(n + zhat^2)) * zhat
s   <- 1
s0  <- n * 0.36
lbf <- z^2/2 * s0/(s0 + s)
pip <- normalizelogweights(lbf)
```

ADD TEXT HERE.

```{r}
j  <- order(pip,decreasing = TRUE)
m  <- which(cumsum(pip[j]) > 0.95)[1]
cs <- j[1:m]
```

ADD TEXT HERE.

```{r, fig.height=1.75, fig.width=5.25}
susie_plot_better(list(pip = pip,sets = list(cs = list(L1 = cs))),causal_snps,
                  xlab = "SNPs on human chromosome 1 near the FMO2 gene")
```

Verify my calculations:

```{r, fig.height=3, fig.width=3}
par(mar = c(4,4,2,1))
fit_1cs <- susie_rss(zhat,R,n = n,L = 1,estimate_prior_method = "EM",
                     tol = 1e-6)
plot(fit_1cs$pip,pip,pch = 20)
abline(a = 0,b = 1,lty = "dotted",col = "magenta")
```

TO DO NEXT: Implement a simple version of the IBSS algorithm.

### References

